---
params:
    tissue: Muscle
    group: 
    - group0
    - group1
    - group2
    p_cutoff: 1e-8
    denom_cutoff: 30
    ir_dist_cutoff: 0.1
title: "LeafcutterMD Analyses - GTEx `r params$tissue`"
author: "Chao Dai"
date: "updated `r Sys.Date()`"

output: 
  bookdown::html_document2:
    toc: yes
    toc_float: no
    theme: readable
    highlight: textmate
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, echo = F)

library(tidyverse)
library(data.table)
library(ggrastr)
library(ggrepel)
library(DT)
library(tictoc)
library(forcats)

theme_set(new = theme_bw())
```

```{r helper-functions, include=FALSE}
# convert a string entry such as "3/4" in to a double "0.75"
# counts file were stored as string of fractions
frac_to_dbl = function(frac_str) {
  frac = str_split(frac_str, "/", simplify=T) %>% as.numeric
  if (frac[[1]] == 0) {
    out = 0
  } else {
    out = frac[[1]] / frac[[2]]
  }
  return(out)
}



get_denom = function(frac_str) {
  # get denominator from a fraction string, eg. get 4 from "3/4"
  frac = str_split(frac_str, "/", simplify=T) %>% as.numeric
  return(frac[[2]])
}
```

<!-- ---------------------------------------------------- -->

# GTEx `r params$tissue` {.tabset}



## `r params$group[[3]]`

```{r}
# load Bladder intron level pvals
tissue = params$tissue
group = params$group[[3]]
p_cutoff = as.numeric(params$p_cutoff)
denom_cutoff = as.numeric(params$denom_cutoff)
ir_dist_cutoff = as.numeric(params$ir_dist_cutoff)
```

```{r}
intron = fread(paste0("../code/results/leafcutterMD/GTEx/", tissue, "/",  group, "_pVals.txt.gz"))
#cluster = fread(paste0("../code/results/leafcutterMD/GTEx/", tissue , "/", group, "_clusterPvals.txt.gz"))
nums = fread(paste0("../code/results/SubsetCounts/GTEx/", tissue, "/nums_filtered/", group, "_perind.nums.gz"))
counts = fread(paste0("../code/results/SubsetCounts/GTEx/", tissue, "/counts/", group, "_perind.counts.gz"),
               header = T)

# rename first column
intron = rename(intron, "chrom" = "V1")
nums = rename(nums, "chrom" = "V1")

# add rowMin
intron_rowMins = select(intron, where(is_numeric)) %>% apply(1, min)
intron$rowMins = intron_rowMins # add min p-val per intron
```

### Intron selection

An intron excision is considered outlier only if: 

  1.  p-value of a given intron excision event \< `r scales::scientific(p_cutoff)`
  2.  **AND** total reads of the cluster \> `r denom_cutoff`
  3. **AND** intron excision ratio is at least `r params$ir_dist_cutoff` away from median

```{r}

# data.table of introns with that have outlier events
# note this is not the final outlier introns yet, as 
# we have to also filter by denominator
outlier_introns = intron[ rowMins < p_cutoff, -c('rowMins')] %>% unique

# get the counts of the same introns
outlier_counts = counts[chrom %in% outlier_introns$chrom]

# melt outlier intron and counts
outlier_introns = melt(outlier_introns, id.vars = "chrom", variable.name = "samples", value.name = "p")
outlier_counts = melt(outlier_counts, id.vars = "chrom", variable.name = "samples", value.name = "frac")

# further filter by denominator
outlier_introns = outlier_introns[outlier_counts, on = c("chrom", "samples"), nomatch = NULL]
```

```{r}
library(furrr)
plan(multisession, workers = availableCores())

denom = future_map_dbl(outlier_introns$frac, get_denom)
ir = future_map_dbl(outlier_introns$frac, frac_to_dbl)

outlier_introns$denom = denom
outlier_introns$ir = ir

```

```{r}
# add in reads
outlier_introns = melt(nums, id.vars = "chrom", variable.name = "samples", value.name = "rc")[
  outlier_introns, on = c("chrom", "samples"), nomatch = NULL
]
```

```{r}
# median IR per intron
outlier_introns[, median_ir := median(ir), by = chrom]
outlier_introns[, dist_ir := abs(median_ir - ir)]

# putative outlier
outlier_introns[, outlier := p < p_cutoff & denom > denom_cutoff & dist_ir > ir_dist_cutoff]

# outlier intron ids
outlier_introns_chroms = outlier_introns[outlier == T, chrom] %>% unique

outlier_introns = outlier_introns[chrom %in% outlier_introns_chroms]
```




### Outlier intron's reads and denom distribution

Some introns have significantly larger amount of total reads (over 10k). May consider set threshold of denominator < 5000. 
```{r}

outlier_introns[outlier == T] %>% 
  ggplot() + stat_ecdf(aes(denom, color = 'denominator') ) +
    stat_ecdf(aes(rc, color = 'numerator')) + 
    scale_x_log10(label = scales::number_format(1)) +
    labs(x = "Outlier introns", y = NULL, color = "measure", caption = "numerator: reads per intron per sample \ndenominator: reads per cluster per sample") +
    theme_bw()

```

Subsampling 100k introns, (all belongs to a cluster that has at least 1 outlier event). Plot suggests no apparent correlation between p value and read counts or denominators. 

```{r}
p = outlier_introns[sample(1:nrow(outlier_introns), 100000)] %>% 
  ggplot() + geom_point(aes( rc,-log10(p), color = 'numerator'), alpha = .2 ) +
    geom_point(aes( denom, -log10(p),color = 'denominator'), alpha = .1 ) + 
    scale_x_log10(label = scales::number_format()) +
    labs(x = 'Subsampled 100k introns', y = '-log10(p)', color = "measure") +
    theme_bw()
rasterize(p, layers = "point", dpi = 50)
```

Take a look at distribution of `sd(ir)`
```{r}
outlier_introns[, .(ir_sd = sd(ir)), by = chrom] %>% 
  ggplot() + stat_ecdf(aes(ir_sd)) + 
    labs(x = "sd of intron excision ratio", y = NULL) + 
    theme_bw() 
```

 Some samples have many more outlier splicing events
```{r, width = 8}
outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples] %>% 
  ggplot()  + geom_col(aes(fct_reorder(samples, -N), N, fill = samples)) + 
    geom_text_repel(aes(fct_reorder(samples, -N), N, label = samples), 
                    data = outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples][N > 500]) + 
    labs(x = NULL) + 
    scale_y_log10() + 
    theme_bw() +
    theme(legend.position = "None", axis.text.x = element_blank())
```
 
```{r}
outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples] %>% 
  ggplot() + stat_ecdf(aes(N)) +
    scale_x_log10() + 
    labs(x = NULL, y = NULL) + 
    theme_bw()
```
```{r}
outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples][, N] %>% quantile(seq(0.5,1, .05))
```



#### Box plot

There are `r outlier_introns[outlier == T, chrom] %>% unique %>% length` outlier introns. 

```{r fig.width=10, fig.height=6, fig.cap="without filters"}

excl_samples = outlier_introns[ outlier == T, .( N = length(unique(chrom))), by = samples][N > 500, samples] %>% as.character
select_chroms = outlier_introns[!samples %in% excl_samples][order(p), chrom] %>% unique %>% head(50)

p2 = outlier_introns[chrom %in% select_chroms] %>% 
  ggplot() + geom_boxplot(aes(fct_reorder(chrom, p, min), ir)) + 
    geom_point(aes(fct_reorder(chrom, p, min), ir, color = outlier), alpha = .4) +
    geom_text_repel(aes(fct_reorder(chrom, p, min), ir, label = samples), 
                    data = outlier_introns[chrom %in% select_chroms & outlier == T & ir == 0]) + 
    labs(x = NULL, y = "Intron excision ratio") + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))

rasterize(p2, layers = "point", dpi=50)
```

```{r fig.width=10, fig.height=6}

excl_samples = outlier_introns[ outlier == T, .( N = length(unique(chrom))), by = samples][N > 500, samples] %>% as.character
select_chroms = outlier_introns[!samples %in% excl_samples][order(p), chrom] %>% unique %>% head(50)

p2 = outlier_introns[chrom %in% select_chroms] %>% 
  ggplot() + geom_boxplot(aes(fct_reorder(chrom, p, min), rc)) + 
    geom_point(aes(fct_reorder(chrom, p, min), rc, color = outlier), alpha = .4) +
    labs(x = NULL, y = "reads") + 
    scale_y_log10() + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))

rasterize(p2, layers = "point", dpi=50)
```

### Outlier introns dug deeper

```{r}

# select low end and high end of ratios
outlier_introns[, .(N_ir_LOW = (ir == 0), N_ir_HIGH = (ir > .05), ir), by = .(chrom, samples)
                ][, .(N_ir_LOW = sum(N_ir_LOW), N_ir_HIGH = sum(N_ir_HIGH)), by = chrom
                  ][order(N_ir_LOW)] %>% 
  ggplot() + geom_point(aes(N_ir_LOW, N_ir_HIGH, color = (N_ir_LOW > 95 & N_ir_HIGH< 5)), alpha =.2)
```

```{r}
outlier_introns[, .(N_ir_LOW = (ir == 0), N_ir_HIGH = (ir > .05), ir), by = .(chrom, samples)
                ][, .(N_ir_LOW = sum(N_ir_LOW), N_ir_HIGH = sum(N_ir_HIGH)), by = chrom
                  ][N_ir_LOW + N_ir_HIGH < 50]
```
```{r}
outlier_introns[chrom %in% "chr15:76225930:76227138:clu_58397_-"]
```


```{r}
select_introns2 = outlier_introns[, .(N_ir_0 = (ir == 0), N_ir_gt5 = (ir > .05), ir), by = .(chrom, samples)
                ][, .(N_ir_0 = sum(N_ir_0), N_ir_gt5 = sum(N_ir_gt5)), by = chrom
                  ][order(-N_ir_0, N_ir_gt5)
                    ][N_ir_0 > 95 & N_ir_gt5 < 4]
select_introns2
```
```{r}
outlier_introns[chrom %in% select_introns2$chrom[1:50]] %>% 
  ggplot() + geom_boxplot(aes(fct_reorder(chrom, p, min), ir)) + 
    geom_point(aes(fct_reorder(chrom, p, min), ir, color = outlier), alpha = .4) +
    geom_text_repel(aes(fct_reorder(chrom, p, min), ir, label = samples), 
                    data = outlier_introns[chrom %in% select_chroms & outlier == T & ir == 0]) + 
    labs(x = NULL, y = "Intron excision ratio") + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))
```





```{r}
outlier_introns[, .(N_ir_1 = (ir > .6), N_ir_lt5 = (ir < .1 & ir > 0), ir), by = .(chrom, samples)
                ][, .(N_ir_1 = sum(N_ir_1), N_ir_lt5 = sum(N_ir_lt5)), by = chrom
                  ][order(-N_ir_1, N_ir_lt5)
                    ][N_ir_1 > 95 & N_ir_lt5 > 0]
```
```{r}
counts[chrom %in% c("chr8:123514334:123531898:clu_199614_-")]
outlier_introns[chrom %in% c("chr8:123514334:123531898:clu_199614_-") & ir < .1]
```


```{r}
counts[chrom %in% c("chr1:223954668:223961168:clu_604817_-", "chr6:32186475:32186604:clu_703502_-")]
outlier_introns[chrom %in% c("chr1:223954668:223961168:clu_604817_-", "chr6:32186475:32186604:clu_703502_-")][
  order(chrom, -ir)
]
```



### QQ plot of intron p-values from 4 randomly selected samples

<!-- by sample -->

```{r  fig.caption = "QQ plot of randomly selected samples"}
# QQ plot of min cluster p-values
pick_4_samples = outlier_introns$samples %>% as.character %>% sample(4)
qq = leafcutter::multiqq(structure(as.list(intron[, ..pick_4_samples]), names = pick_4_samples))
rasterize(qq, layers = 'point', dpi = 50)
```

<!-- qq plot of selected introns (not shown) -->

```{r fig.cap="QQ plot of introns with outlier splicing", include=FALSE, eval=FALSE}
intron[cluster %in% outlier_clusters & rowMins < p_cutoff][
  order(rowMins)][1:1000,] %>% 
  .[, -c("cluster", "rowMins")] %>% 
  column_to_rownames("chrom") %>% t %>% as.data.frame %>% as.list %>% 
  leafcutter::multiqq(.) + theme_bw() + 
    guides(color = guide_legend(nrow=2, title=NULL)) + 
    theme(legend.position = "bottom")
```





### high outlier samples

```{r}
samples_set1 = outlier_introns[outlier == T][
  , .(N = length(unique(chrom))), by = samples][
    N > 500, samples
  ] %>% as.character

samples_set2 = outlier_introns[outlier == T][
  , .(N = length(unique(chrom))), by = samples][
    N < 100, samples
  ] %>% as.character

GTEx = fread("../code/resources/snaptron-yil/scripts/data/samples_gtex.tsv")

```


Samples with > 500 outliers

```{r}
GTEx[rail_id %in% samples_set1][
  , .(rail_id, Sex, AGE, Histological_Type, Analyte_Type, SMPTHNTS, SMSMPSTE, SMSTYP, SMNABTCHT, TRTPTREF, DTHFUCOD, DTHMNNR)
]

```


Samples with less than 100 outliers
```{r}

GTEx[rail_id %in% samples_set2][
  , .(rail_id, Sex, AGE, Histological_Type, Analyte_Type, SMPTHNTS, SMSMPSTE, SMSTYP, SMNABTCHT, TRTPTREF, DTHFUCOD, DTHMNNR)
] %>% head
```

```{r include=FALSE}
rm(intron, counts, denom, outlier_introns, nums)
gc()
```

------


## `r params$group[[2]]`

```{r}
# load Bladder intron level pvals
tissue = params$tissue
group = params$group[[2]]
p_cutoff = as.numeric(params$p_cutoff)
denom_cutoff = as.numeric(params$denom_cutoff)
ir_dist_cutoff = as.numeric(params$ir_dist_cutoff)
```

```{r}
intron = fread(paste0("../code/results/leafcutterMD/GTEx/", tissue, "/",  group, "_pVals.txt.gz"))
#cluster = fread(paste0("../code/results/leafcutterMD/GTEx/", tissue , "/", group, "_clusterPvals.txt.gz"))
nums = fread(paste0("../code/results/SubsetCounts/GTEx/", tissue, "/nums_filtered/", group, "_perind.nums.gz"))
counts = fread(paste0("../code/results/SubsetCounts/GTEx/", tissue, "/counts/", group, "_perind.counts.gz"),
               header = T)

# rename first column
intron = rename(intron, "chrom" = "V1")
nums = rename(nums, "chrom" = "V1")

# add rowMin
intron_rowMins = select(intron, where(is_numeric)) %>% apply(1, min)
intron$rowMins = intron_rowMins # add min p-val per intron
```

### Intron selection

An intron excision is considered outlier only if: 

  1.  p-value of a given intron excision event \< `r scales::scientific(p_cutoff)`
  2.  **AND** total reads of the cluster \> `r denom_cutoff`
  3. **AND** intron excision ratio is at least `r params$ir_dist_cutoff` away from median

```{r}

# data.table of introns with that have outlier events
# note this is not the final outlier introns yet, as 
# we have to also filter by denominator
outlier_introns = intron[ rowMins < p_cutoff, -c('rowMins')] %>% unique

# get the counts of the same introns
outlier_counts = counts[chrom %in% outlier_introns$chrom]

# melt outlier intron and counts
outlier_introns = melt(outlier_introns, id.vars = "chrom", variable.name = "samples", value.name = "p")
outlier_counts = melt(outlier_counts, id.vars = "chrom", variable.name = "samples", value.name = "frac")

# further filter by denominator
outlier_introns = outlier_introns[outlier_counts, on = c("chrom", "samples"), nomatch = NULL]
```

```{r}
library(furrr)
plan(multisession, workers = availableCores())

denom = future_map_dbl(outlier_introns$frac, get_denom)
ir = future_map_dbl(outlier_introns$frac, frac_to_dbl)

outlier_introns$denom = denom
outlier_introns$ir = ir

```

```{r}
# add in reads
outlier_introns = melt(nums, id.vars = "chrom", variable.name = "samples", value.name = "rc")[
  outlier_introns, on = c("chrom", "samples"), nomatch = NULL
]
```

```{r}
# median IR per intron
outlier_introns[, median_ir := median(ir), by = chrom]
outlier_introns[, dist_ir := abs(median_ir - ir)]

# putative outlier
outlier_introns[, outlier := p < p_cutoff & denom > denom_cutoff & dist_ir > ir_dist_cutoff]

# outlier intron ids
outlier_introns_chroms = outlier_introns[outlier == T, chrom] %>% unique

outlier_introns = outlier_introns[chrom %in% outlier_introns_chroms]
```




### Outlier intron's reads and denom distribution

Some introns have significantly larger amount of total reads (over 10k). May consider set threshold of denominator < 5000. 
```{r}

outlier_introns[outlier == T] %>% 
  ggplot() + stat_ecdf(aes(denom, color = 'denominator') ) +
    stat_ecdf(aes(rc, color = 'numerator')) + 
    scale_x_log10(label = scales::number_format(1)) +
    labs(x = "Outlier introns", y = NULL, color = "measure", caption = "numerator: reads per intron per sample \ndenominator: reads per cluster per sample") +
    theme_bw()

```

Subsampling 100k introns, (all belongs to a cluster that has at least 1 outlier event). Plot suggests no apparent correlation between p value and read counts or denominators. 

```{r}
p = outlier_introns[sample(1:nrow(outlier_introns), 100000)] %>% 
  ggplot() + geom_point(aes( rc,-log10(p), color = 'numerator'), alpha = .2 ) +
    geom_point(aes( denom, -log10(p),color = 'denominator'), alpha = .1 ) + 
    scale_x_log10(label = scales::number_format()) +
    labs(x = 'Subsampled 100k introns', y = '-log10(p)', color = "measure") +
    theme_bw()
rasterize(p, layers = "point", dpi = 50)
```

Take a look at distribution of `sd(ir)`
```{r}
outlier_introns[, .(ir_sd = sd(ir)), by = chrom] %>% 
  ggplot() + stat_ecdf(aes(ir_sd)) + 
    labs(x = "sd of intron excision ratio", y = NULL) + 
    theme_bw() 
```

 Some samples have many more outlier splicing events
```{r, width = 8}
outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples] %>% 
  ggplot()  + geom_col(aes(fct_reorder(samples, -N), N, fill = samples)) + 
    geom_text_repel(aes(fct_reorder(samples, -N), N, label = samples), 
                    data = outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples][N > 500]) + 
    labs(x = NULL) + 
    scale_y_log10() + 
    theme_bw() +
    theme(legend.position = "None", axis.text.x = element_blank())
```
 
```{r}
outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples] %>% 
  ggplot() + stat_ecdf(aes(N)) +
    scale_x_log10() + 
    labs(x = NULL, y = NULL) + 
    theme_bw()
```
```{r}
outlier_introns[outlier == T][, .(N = length(unique(chrom))), by = samples][, N] %>% quantile(seq(0.5,1, .05))
```



#### Box plot

There are `r outlier_introns[outlier == T, chrom] %>% unique %>% length` outlier introns. 

```{r fig.width=10, fig.height=6, fig.cap="without filters"}

excl_samples = outlier_introns[ outlier == T, .( N = length(unique(chrom))), by = samples][N > 500, samples] %>% as.character
select_chroms = outlier_introns[!samples %in% excl_samples][order(p), chrom] %>% unique %>% head(50)

p2 = outlier_introns[chrom %in% select_chroms] %>% 
  ggplot() + geom_boxplot(aes(fct_reorder(chrom, p, min), ir)) + 
    geom_point(aes(fct_reorder(chrom, p, min), ir, color = outlier), alpha = .4) +
    geom_text_repel(aes(fct_reorder(chrom, p, min), ir, label = samples), 
                    data = outlier_introns[chrom %in% select_chroms & outlier == T & ir == 0]) + 
    labs(x = NULL, y = "Intron excision ratio") + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))

rasterize(p2, layers = "point", dpi=50)
```

```{r fig.width=10, fig.height=6}

excl_samples = outlier_introns[ outlier == T, .( N = length(unique(chrom))), by = samples][N > 500, samples] %>% as.character
select_chroms = outlier_introns[!samples %in% excl_samples][order(p), chrom] %>% unique %>% head(50)

p2 = outlier_introns[chrom %in% select_chroms] %>% 
  ggplot() + geom_boxplot(aes(fct_reorder(chrom, p, min), rc)) + 
    geom_point(aes(fct_reorder(chrom, p, min), rc, color = outlier), alpha = .4) +
    labs(x = NULL, y = "reads") + 
    scale_y_log10() + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))

rasterize(p2, layers = "point", dpi=50)
```




### QQ plot of intron p-values from 4 randomly selected samples

<!-- by sample -->

```{r  fig.caption = "QQ plot of randomly selected samples"}
# QQ plot of min cluster p-values
pick_4_samples = outlier_introns$samples %>% as.character %>% sample(4)
qq = leafcutter::multiqq(structure(as.list(intron[, ..pick_4_samples]), names = pick_4_samples))
rasterize(qq, layers = 'point', dpi = 50)
```

<!-- qq plot of selected introns (not shown) -->

```{r fig.cap="QQ plot of introns with outlier splicing", include=FALSE, eval=FALSE}
intron[cluster %in% outlier_clusters & rowMins < p_cutoff][
  order(rowMins)][1:1000,] %>% 
  .[, -c("cluster", "rowMins")] %>% 
  column_to_rownames("chrom") %>% t %>% as.data.frame %>% as.list %>% 
  leafcutter::multiqq(.) + theme_bw() + 
    guides(color = guide_legend(nrow=2, title=NULL)) + 
    theme(legend.position = "bottom")
```





### high outlier samples

```{r}
samples_set1 = outlier_introns[outlier == T][
  , .(N = length(unique(chrom))), by = samples][
    N > 500, samples
  ] %>% as.character

samples_set2 = outlier_introns[outlier == T][
  , .(N = length(unique(chrom))), by = samples][
    N < 100, samples
  ] %>% as.character

GTEx = fread("../code/resources/snaptron-yil/scripts/data/samples_gtex.tsv")

```


Samples with > 500 outliers

```{r}
GTEx[rail_id %in% samples_set1][
  , .(rail_id, Sex, AGE, Histological_Type, Analyte_Type, SMPTHNTS, SMSMPSTE, SMSTYP, SMNABTCHT, TRTPTREF, DTHFUCOD, DTHMNNR)
]

```


Samples with less than 100 outliers
```{r}

GTEx[rail_id %in% samples_set2][
  , .(rail_id, Sex, AGE, Histological_Type, Analyte_Type, SMPTHNTS, SMSMPSTE, SMSTYP, SMNABTCHT, TRTPTREF, DTHFUCOD, DTHMNNR)
] %>% head
```

```{r include=FALSE}
rm(intron, counts, denom, outlier_introns, nums)
gc()
```
