---
title: "leafcutterMD"
author: "daiuc"
date: "2022-07-25"
output: 
  workflowr::wflow_html:
    code_folding: "hide"
    df_print: "paged"
  
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
```

# Gathered Samples

```{r}
all_perind_lookup = fread("../code/resources/All_perind_rail_id_tissue_type_lookup.tsv")

all_perind_lookup[ds %in% c("GTEx", "TCGA"),][
  , .(.N), by = .(tissue, ds)
][order(N)]
```

Total number of samples to be analyzed: 28881, a few hundred less than the count table, due to insufficient tissue data available. 

```{r}
length(all_perind_lookup$rail_id)
```



# LeafcutterMD output examples

Processing procedures: 

  1. produce lookup table (rail_id : [tissue, data source (GTEx, TCGA, SRAV)])
  2. subset count table by tissue and datasource
  3. prep leafcutterMD input count matrix
    - necessary header/numerator data wrangling
    - filter intron clusters with less than 5 total reads
  4. leafcutterMD


```{r, fig.align='center'}
knitr::include_graphics("assets/leafcutterMD-dag.svg", error = F)
```




